A simple and efficient algorithm for computing approximate Nash equilibria in two-person zero-sum sequential games with imcomplete information

### 2015-07-04 nips2015.tex
fixups

\paragraph{\textbf{Convex analysis.}}
Given a subset $C$ of $\mathbb{R}^n$,
$i_C$ denotes its \textit{indicator function} defined by
\begin{eqnarray}
  i_C(x) = 0 \text{ if } x \in C\text{ and }+\infty\text{
    otherwise.}
\end{eqnarray}
At times, we will write $i_{x \in C}$ for $i_C(x)$ (to ease notation,
etc.). Let $f : \mathbb{R}^n \rightarrow (-\infty, +\infty]$ be a
  convex function. The \textit{effective domain} of $f$, denoted
  $dom(f)$ is defined as
\begin{eqnarray}
  dom(f) := \{x \in \mathbb{R}^n | f(x) < +\infty\}.
\end{eqnarray}
 If $dom(f) \ne \emptyset$ then we say $f$ is \textit{proper}.
%% \textit{proper convex lower semi-continous function}
%% (\textit{p.c.l.s.c} for short).
%% The \textit{Fenchel-Legendre transform} of $f$ is the function $f^*:
%% \mathbb{R}^n \rightarrow (-\infty, +\infty]$ defined by $f^*(x) \equiv
%%   \underset{z \in \mathbb{R}^n}{\text{max}}\text{}z^Tx - f(z)$.
The \textit{subgradient} of $f$ is the set-valued function
\begin{eqnarray}
\partial f: \mathbb{R}^n \rightarrow 2^{\mathbb{R}^n}, \hspace{.5em} x \mapsto
\{s\in \mathbb{R}^n | f(z)  \ge f(x) + \langle s, z - x\rangle, \forall
z \in \mathbb{R}^n\}.
\end{eqnarray}
Of course $\partial f(x)$ reduces to the singleton $\{\nabla f(x)\}$
in case $f$ is differentiable at $x$. If $f$ is convex, its
\textit{proximal operator} the function defined by
\begin{eqnarray}
  \prox_f: \mathbb{R}^n \rightarrow \mathbb{R}^n,\hspace{.5em}x
  \mapsto \underset{z \in \mathbb{R}^n}{\text{min }}\frac{1}{2}\|z
  - x\|^2 + f(z),
\end{eqnarray}
For example, if $C$ is a closed convex subset of $\mathbb{R}^n$, then
$\prox_C = \proj_C$, the orthogonal projector onto $C$. Thus proximal
operators generalize orthongal projectors. One also has the useful
characterization
\begin{eqnarray}
  p = \prox_f(x)\text{ iff } x - p \in \partial f(p).
\end{eqnarray}

@@

and the dual LCP
\begin{equation}
  \begin{aligned}
    \text{maximize}-\langle e_2, q\rangle,\hspace{.5em}\text{subject to: }(x,q) \in
    \mathbb{R}^{n_1} \times \mathbb{R}^{l_2},  x \ge 0, E_1x = e_1, A^Tx + E_2^Tq \geq 0.
  \end{aligned}
  \label{eq:dual_pb}
\end{equation}
The vectors $p = (p_0, p_1, ..., p_{l_2 - 1}) \in \mathbb{R}^{l_2}$
and $q = (q_0, q_1, ..., q_{l_1 - 1}) \in \mathbb{R}^{l_1}$ are dual
variables. 
$A$ is the \textit{payoff matrix} and each $E_k$ is a matrix whose
entries are $-1$, $0$ or $1$, and each $e_k$ is a vector of the form
$(1, 0, ..., 0)$. In this so-called \textit{sequence-form}
representation, the strategy profile of player $k$ is the polyhedron
\begin{equation}
  Q_k := \{z \in \mathbb{R}^{n_k}_+ |\text{ }E_kz = e_k\}.
\label{eq:polyhedron}
\end{equation}
%% Note that $\langle e_1, p\rangle = p_0$ and $\langle e_2, q\rangle = q_0$ by definition of $e_1$ and
%% $e_2$. 
At a feasible point $(y, p, x, q)$, the \textit{primal-dual gap}
$\mathcal{G}(y, p, x, q)$ of these primal-dual pair of LCPs is given
by\footnote{The inequality being due to \textit{weak duality}.}
\begin{equation}
  0 \le \mathcal{G}(y, p, x, q) := \langle e_1, p\rangle - (\langle
  e_2, q\rangle) = \langle e_1, p\rangle + \langle
  e_2, q\rangle = G(x, y),
  \label{eq:dgap}
\end{equation}
where
\begin{eqnarray}
G(x, y) :=
\mathrm{max}\{\langle u, Ay\rangle - \langle x, Av\rangle |
(u,v) \in Q_1 \times Q_2\}.
\label{eq:xy_gap}
\end{eqnarray}

@@

\begin{remark}  
It should be noted that for the the class of games considered here
(sequential games with imcomplete information), the LCPs
\eqref{eq:primal_pb} and \eqref{eq:dual_pb} are exceedingly larger
than what state-of-the-art LCP solvers can handle. See
\cite{hoda2010smoothing}.

@@

In \cite{hoda2010smoothing}, a nested iterative procedure using the
Excessive Gap Technique (EGT) \cite{nesterov2005excessive} was used to
to solve the LCPs \eqref{eq:primal_pb} and \eqref{eq:dual_pb} in the
following equivalent saddle-point form\footnote{Note that in
  \cite{hoda2010smoothing}, the roles of $x$ and $y$ are swapped.}

\begin{equation}
  \underset{y \in Q_2}{\text{minimize}}\text{ }\underset{x \in
    Q_1}{\text{maximize}}\text{ }\langle x, Ay\rangle
  \label{eq:gilpin}
\end{equation}
The authors reported a $\mathcal{O}(1/\epsilon)$ convergence rate
(which derives from the general EGT theory) for the outer-most
iteration loop.

\cite{gilpinfirst} proposed a modified verson of the technqiues in
\cite{hoda2010smoothing} and  proved a $\mathcal{O}\left(\left(\|A\| /
\delta\right) ln\left(1 / \epsilon\right)\right)$ convergence rate in
terms of the number of calls made to a first-order oracle. Here
$\delta = \delta(A, E_1, E_2, e_1, e_2) > 0$ is a certain
\textit{condition number} for the game. The crux of their technique was to
observe that \eqref{eq:gilpin} can further be written a the minimization of
the primal-dual gap function $G(x, y)$ (defined in \eqref{eq:xy_gap})
for the game\footnote{The minimizers of $G$ are precisely the
  equilibria of the game.}, viz
\begin{eqnarray}
\mathrm{maximize}\{G(x,y)|(x,y) \in Q_1 \times Q_2\},
\end{eqnarray}
and then show there exists a scalar
$\delta > 0$ such that for any pair of realization plans $(x, y) \in Q_1 \times Q_2$,
\begin{eqnarray}
\text{``distance between }(x, y)\text{ and the set of
equilibria'' } \le G(x, y)/\delta.
\end{eqnarray}
Their
algorithm is then derived by iteratively applying Nesterov smoothing \cite{nesterov2005a}
with a geometrically decreasing sequence of tolerance levels
$\epsilon_{n+1} = \epsilon_n / \gamma$ (with $\gamma > 1$)  on
primal-dual gap function $G$. It should be noted however that
\begin{itemize}
\item[--] The constant $\delta > 0$ can be arbitrarily small, and so
  the factor $\|A\| / \delta$ in the $\mathcal{O}\left(\left(\|A\| /
\delta\right) ln\left(1 / \epsilon\right)\right)$ convergence rate can
be arbitrarily large.
\item[--] The reported linear convergence rate is not in terms of
  basic operations (addition, multiplication, matvec, clipping, etc.),
  but in terms of the number of calls to a first-order oracle. For
  example the projections $\proj_{Q_k}$ are applied at each iteration.%%  Of
  %% course, the existence of a linear algorithm for solving
  %% \eqref{eq:gilpin} is improbable, as the usual strong-convexity
  %% conditions are absent.

@@

Also define two bilinear forms $\Psi_1$, $\Psi_2: \mathbb{R}^{n_2}
\times \mathbb{R}^{l_1} \times \mathbb{R}^{n_1} \times
\mathbb{R}^{l_2} \rightarrow \mathbb{R}$ with $\Psi_2 = -\Psi_1$ by letting
  \begin{equation}
      K :=
      \left[
        \begin{array}{cc}
          A & -E_1^T \\
          E_2 & 0
        \end{array}
        \right],\hspace{.5em}
    \Psi_1(y, p, x, q)
    := \left\langle \begin{bmatrix}x\\q\end{bmatrix},
      K\begin{bmatrix}y\\p\end{bmatrix}\right\rangle.
\end{equation}

@@

\begin{definition}(Kronecker product)
Let $F \in \mathbb{R}^{m \times n}$ and $B \in \mathbb{R}^{r \times
  s}$ be matrices. The kronecker product of $F$ and $B$ is defined by

\begin{equation}
F \otimes B:=\left[
\begin{array}{cccc}
F_{0,0}B & F_{0,1}B & \cdots & F_{0,n-1}B \\
F_{1,0}B & F_{1,1}B & \cdots & F_{1,n-1}B \\
\vdots & \vdots & \ddots & \vdots\\
F_{m-1,0}B & F_{m-1,1}B & \cdots & F_{m-1,n-1}B 
\end{array}\right] \in \mathbb{R}^{mr \times ns}
\end{equation}
\end{definition}

One notes that $(F \otimes B)^T = F^T \otimes B^T$.
$F \otimes B$ can be very much larger than both $F$ and $B$ in
dimensions. For example, if $r = n = r = s = 1000$, so that $F$ and
$B$ are $10^3$-by-$10^3$ matrices, then $F \otimes B$ is a
$10^6$-by-$10^6$ matrix! Fortunately, given a vector $x \in
\mathbb{R}^{rs}$ one can compute the matvec product $(F \otimes B)x$
without forming $F \otimes B$. Indeed, let $y := (F \otimes B)x$. We
rewrite the row vector $x$ as an $s$-by-$r$ matrix $X = [X_0, X_1,
  ..., X_{r-1}]$, where each column vector $X_i$ is the $i$th block of
$s$ elements of $x$ read from left to right. Similarly, rewrite $y$ as
as an $n$-by-$m$ matrix $Y = [Y_0, Y_1, ..., Y_{m-1}]$. Then one
easily verfies that $Y$ is the matrix product of the ``small''
matrices $B$, $X$, and $F^T$, i.e

\begin{equation}
  \label{eq:kron_matvec}
  Y = BXF^T
\end{equation}

Now, in Texas Hold'em for example, the payoff matrix $A$ can be
written as block-diagonal matrix whose blocks are sums of Kronecker
products of much smaller sparse matrices as follows

\begin{equation}
  A = \begin{bmatrix}F_1 \otimes B_1\hspace{10em}\\\hspace{3em}F_2
    \otimes B_2\hspace{7em}\\\hspace{6em}F_3 \otimes
    B_3\hspace{4em}\\\hspace{10em}F_4 \otimes B_4 + S \end{bmatrix}
\label{eq:factor_A}
\end{equation}
See \cite{hoda2010smoothing}. The matrices $F_i$ correspond to
sequences of moves in round $i$ which end in a \textit{fold} action,
and $S$ to the sequences which end in a \textit{showdown}. $B_i$
encodes the betting structure of round $i$, while $W$ encodes the
wind/lose/draw information determined by ranking the players' hands at
showdown. The component matrices $F_i$, $B_i$, $S$, and $W$ are small
enough to be explictly represented whereas it is infeasible to
explicitly represent A. Furthermore, the matrices $F_i$, $B_i$,
$S$, and $W$ are themselves sparse, which allows one to use the
compressed row storage data structure that only stores nonzero entries
(for example \textit{scipy.sparse.csr\_matrix}, in the Python
programming language).

Such a representation of the payoff matrix $A$ trivializes matvec
operations involving $A$ or $A^T$ (thanks to formula
\eqref{eq:kron_matvec} above, applied to their diagonal blocks and
exploiting the sparsity of these blocks themselves). Of course, one
can always write the payoff matrix $A$ of a flop-type Poker game in a
form similary to \eqref{eq:factor_A} by applying appropriate
permutations to the enumeration of the players's sequences.

>>>>

\paragraph{\textbf{Convex analysis.}}
Given a subset $C$ of $\mathbb{R}^n$,
$i_C$ denotes its \textit{indicator function} defined by
\begin{eqnarray}
  i_C(x) = 0 \text{ if } x \in C\text{ and }+\infty\text{
    otherwise.}
\end{eqnarray}
At times, we will write $i_{x \in C}$ for $i_C(x)$ (to ease notation,
etc.). Let $f : \mathbb{R}^n \rightarrow (-\infty, +\infty]$ be a
  convex function. The \textit{effective domain} of $f$, denoted
  $dom(f)$ is defined as
\begin{eqnarray}
  dom(f) := \{x \in \mathbb{R}^n | f(x) < +\infty\}.
\end{eqnarray}
 If $dom(f) \ne \emptyset$ then we say $f$ is \textit{proper}.
%% \textit{proper convex lower semi-continous function}
%% (\textit{p.c.l.s.c} for short).
%% The \textit{Fenchel-Legendre transform} of $f$ is the function $f^*:
%% \mathbb{R}^n \rightarrow (-\infty, +\infty]$ defined by $f^*(x) \equiv
%%   \underset{z \in \mathbb{R}^n}{\text{max}}\text{}z^Tx - f(z)$.
The \textit{subgradient} of $f$ is the set-valued function
\begin{eqnarray}
\partial f: \mathbb{R}^n \rightarrow 2^{\mathbb{R}^n}, \hspace{.5em} x \mapsto
\{s\in \mathbb{R}^n | f(z)  \ge f(x) + \langle s, z - x\rangle, \forall
z \in \mathbb{R}^n\}.
\end{eqnarray}
Of course $\partial f(x)$ reduces to the singleton $\{\nabla f(x)\}$
in case $f$ is differentiable at $x$. If $f$ is convex, its
\textit{proximal operator} the function defined by
\begin{eqnarray}
  \prox_f: \mathbb{R}^n \rightarrow \mathbb{R}^n,\hspace{.5em}x
  \mapsto \underset{z \in \mathbb{R}^n}{\text{min }}\frac{1}{2}\|z
  - x\|^2 + f(z),
\end{eqnarray}
For example, if $C$ is a closed convex subset of $\mathbb{R}^n$, then
$\prox_C = \proj_C$, the orthogonal projector onto $C$. Thus proximal
operators generalize orthogonal projectors. One also has the useful
characterization
\begin{eqnarray}
  p = \prox_f(x)\text{ iff } x - p \in \partial f(p).
\end{eqnarray}

@@

and the dual LCP
\begin{equation}
  \begin{aligned}
    \text{maximize}-\langle e_2, q\rangle,\hspace{.5em}\text{subject
      to: }(x,q) \in \mathbb{R}^{n_1} \times \mathbb{R}^{l_2},  x \ge
    0, E_1x = e_1, A^Tx + E_2^Tq \geq 0.
  \end{aligned}
  \label{eq:dual_pb}
\end{equation}
The vectors $p = (p_0, p_1, ..., p_{l_2 - 1}) \in \mathbb{R}^{l_2}$
and $q = (q_0, q_1, ..., q_{l_1 - 1}) \in \mathbb{R}^{l_1}$ are dual
variables. 
$A$ is the \textit{payoff matrix} and each $E_k$ is a matrix whose
entries are $-1$, $0$ or $1$, and each $e_k$ is a vector of the form
$(1, 0, ..., 0)$. In this so-called \textit{sequence-form}
representation, the strategy profile of player $k$ is the polyhedron
\begin{equation}
  Q_k := \{z \in \mathbb{R}^{n_k}_+ |\text{ }E_kz = e_k\}.
\label{eq:polyhedron}
\end{equation}

Note that the LCPs above have the equivalent saddle point formulation
\begin{equation}
  \underset{y \in Q_2}{\text{minimize}}\text{ }\underset{x \in
    Q_1}{\text{maximize}}\text{ }\langle x, Ay\rangle
  \label{eq:gilpin}
\end{equation}

%% Note that $\langle e_1, p\rangle = p_0$ and $\langle e_2, q\rangle
%% = q_0$ by definition of $e_1$ and
%% $e_2$. 
At a feasible point $(y, p, x, q)$, the \textit{primal-dual gap}
$\mathcal{G}(y, p, x, q)$ of these primal-dual pair of LCPs is given
by\footnote{The inequality being due to \textit{weak duality}.}
\begin{equation}
  0 \le \mathcal{G}(y, p, x, q) := \langle e_1, p\rangle - (\langle
  e_2, q\rangle) = \langle e_1, p\rangle + \langle
  e_2, q\rangle = G(x, y),
  \label{eq:dgap}
\end{equation}
where
\begin{eqnarray}
G(x, y) :=
\mathrm{max}\{\langle u, Ay\rangle - \langle x, Av\rangle |
(u,v) \in Q_1 \times Q_2\}.
\label{eq:xy_gap}
\end{eqnarray}

@@

\begin{remark}  
It should be noted that for the class of games considered here
(sequential games with imcomplete information), the LCPs
\eqref{eq:primal_pb} and \eqref{eq:dual_pb} are exceedingly larger
than what state-of-the-art LCP solvers can handle. See
\cite{hoda2010smoothing}.

@@

In \cite{hoda2010smoothing}, a nested iterative procedure using the
Excessive Gap Technique (EGT) \cite{nesterov2005excessive} was used to
to solve the equilibrium problem \eqref{eq:gilpin}.
The authors reported a $\mathcal{O}(1/\epsilon)$ convergence rate
(which derives from the general EGT theory) for the outer-most
iteration loop.

\cite{gilpinfirst} proposed a modified verson of the technqiues in
\cite{hoda2010smoothing} and  proved a $\mathcal{O}\left(\left(\|A\| /
\delta\right) ln\left(1 / \epsilon\right)\right)$ convergence rate in
terms of the number of calls made to a first-order oracle. Here
$\delta = \delta(A, E_1, E_2, e_1, e_2) > 0$ is a certain
\textit{condition number} for the game. The crux of their technique was to
observe that \eqref{eq:gilpin} can further be written a the minimization of
the primal-dual gap function $G(x, y)$ (defined in \eqref{eq:xy_gap})
for the game\footnote{The minimizers of $G$ are precisely the
  equilibria of the game.}, viz
\begin{eqnarray}
\mathrm{maximize}\{G(x,y)|(x,y) \in Q_1 \times Q_2\},
\end{eqnarray}
and then show there exists a scalar
$\delta > 0$ such that for any pair of realization plans $(x, y) \in Q_1 \times Q_2$,
\begin{eqnarray}
\text{``distance between }(x, y)\text{ and the set of
equilibria'' } \le G(x, y)/\delta.
\end{eqnarray}
Their
algorithm is then derived by iteratively applying Nesterov smoothing \cite{nesterov2005a}
with a geometrically decreasing sequence of tolerance levels
$\epsilon_{n+1} = \epsilon_n / \gamma$ (with $\gamma > 1$)  $G$. It
should be noted however that
\begin{itemize}
\item[--] The constant $\delta > 0$ can be arbitrarily small, and so
  the factor $\|A\| / \delta$ in the $\mathcal{O}\left(\left(\|A\| /
\delta\right) ln\left(1 / \epsilon\right)\right)$ convergence rate can
be arbitrarily large.
\item[--] The reported linear convergence rate is not in terms of
  basic operations (addition, multiplication, matvec, clipping, etc.),
  but in terms of the number of calls to a first-order oracle. For
  example the projections $\proj_{Q_k}$ are applied at each iteration.%%  Of
  %% course, the existence of a linear algorithm for solving
  %% \eqref{eq:gilpin} is improbable, as the usual strong-convexity
  %% conditions are absent.

@@

Also define two bilinear forms $\Psi_1$, $\Psi_2: \mathbb{R}^{n_2}
\times \mathbb{R}^{l_1} \times \mathbb{R}^{n_1} \times
\mathbb{R}^{l_2} \rightarrow \mathbb{R}$ with $\Psi_2 = -\Psi_1$ by letting
  \begin{equation}
      K :=
      \left[
        \begin{array}{cc}
          A & -E_1^T \\
          E_2 & 0
        \end{array}
        \right],\hspace{.5em}
    \Psi_1(y, p, x, q)
    := \left\langle \begin{bmatrix}x\\q\end{bmatrix},
      K\begin{bmatrix}y\\p\end{bmatrix}\right\rangle = \langle x,
      Ay\rangle -\langle x, E_1^Tp\rangle + \langle q, E_2y\rangle.
\end{equation}

@@

%% \begin{definition}(Kronecker product)
%% Let $F \in \mathbb{R}^{m \times n}$ and $B \in \mathbb{R}^{r \times
%%   s}$ be matrices. The kronecker product of $F$ and $B$ is defined by

%% \begin{equation}
%% F \otimes B:=\left[
%% \begin{array}{cccc}
%% F_{0,0}B & F_{0,1}B & \cdots & F_{0,n-1}B \\
%% F_{1,0}B & F_{1,1}B & \cdots & F_{1,n-1}B \\
%% \vdots & \vdots & \ddots & \vdots\\
%% F_{m-1,0}B & F_{m-1,1}B & \cdots & F_{m-1,n-1}B 
%% \end{array}\right] \in \mathbb{R}^{mr \times ns}
%% \end{equation}
%% \end{definition}

%% One notes that $(F \otimes B)^T = F^T \otimes B^T$.
%% $F \otimes B$ can be very much larger than both $F$ and $B$ in
%% dimensions. For example, if $r = n = r = s = 1000$, so that $F$ and
%% $B$ are $10^3$-by-$10^3$ matrices, then $F \otimes B$ is a
%% $10^6$-by-$10^6$ matrix! Fortunately, given a vector $x \in
%% \mathbb{R}^{rs}$ one can compute the matvec product $(F \otimes B)x$
%% without forming $F \otimes B$. Indeed, let $y := (F \otimes B)x$. We
%% rewrite the row vector $x$ as an $s$-by-$r$ matrix $X = [X_0, X_1,
%%   ..., X_{r-1}]$, where each column vector $X_i$ is the $i$th block of
%% $s$ elements of $x$ read from left to right. Similarly, rewrite $y$ as
%% as an $n$-by-$m$ matrix $Y = [Y_0, Y_1, ..., Y_{m-1}]$. Then one
%% easily verfies that $Y$ is the matrix product of the ``small''
%% matrices $B$, $X$, and $F^T$, i.e

%% \begin{equation}
%%   \label{eq:kron_matvec}
%%   Y = BXF^T
%% \end{equation}

%% Now, in Texas Hold'em for example, the payoff matrix $A$ can be
%% written as block-diagonal matrix whose blocks are sums of Kronecker
%% products of much smaller sparse matrices as follows

%% \begin{equation}
%%   A = \begin{bmatrix}F_1 \otimes B_1\hspace{10em}\\\hspace{3em}F_2
%%     \otimes B_2\hspace{7em}\\\hspace{6em}F_3 \otimes
%%     B_3\hspace{4em}\\\hspace{10em}F_4 \otimes B_4 + S \end{bmatrix}
%% \label{eq:factor_A}
%% \end{equation}
%% See \cite{hoda2010smoothing}. The matrices $F_i$ correspond to
%% sequences of moves in round $i$ which end in a \textit{fold} action,
%% and $S$ to the sequences which end in a \textit{showdown}. $B_i$
%% encodes the betting structure of round $i$, while $W$ encodes the
%% wind/lose/draw information determined by ranking the players' hands at
%% showdown. The component matrices $F_i$, $B_i$, $S$, and $W$ are small
%% enough to be explictly represented whereas it is infeasible to
%% explicitly represent A. Furthermore, the matrices $F_i$, $B_i$,
%% $S$, and $W$ are themselves sparse, which allows one to use the
%% compressed row storage data structure that only stores nonzero entries
%% (for example \textit{scipy.sparse.csr\_matrix}, in the Python
%% programming language).

%% Such a representation of the payoff matrix $A$ trivializes matvec
%% operations involving $A$ or $A^T$ (thanks to formula
%% \eqref{eq:kron_matvec} above, applied to their diagonal blocks and
%% exploiting the sparsity of these blocks themselves). Of course, one
%% can always write the payoff matrix $A$ of a flop-type Poker game in a
%% form similary to \eqref{eq:factor_A} by applying appropriate
%% permutations to the enumeration of the players's sequences.
